---
title: "Voter Data Analysis and Visualization from the 2022 federal election"
author: "Tuguldur Ganbat"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE, eval = TRUE)
library(tidyverse)
library(sf)
library(here)
```

**Q1) (1 mark) Based on the voter data from the 2022 federal election, which electorates were won by Independent candidates?**

```{r}
# Step 1: Read the CSV data from the provided URL, skipping the first row
election_data <- read_csv("https://results.aec.gov.au/27966/Website/Downloads/HouseDopByDivisionDownload-27966.csv", skip = 1)

# Step 2: Filter the data
election_data <- election_data |>
  filter(
    CalculationType == "Preference Count" &  
    Elected == "Y" &  
    CountNumber == 0 &  
    PartyNm == "Independent"  
  ) |>
  mutate(DivisionNm = toupper(DivisionNm))  

# Step 3: Select DivisionNm column
election_data %>% 
  select(DivisionNm)

```

**Q2) (1 mark) How many of these were formerly liberal seats that were lost to independents in 2022?**

```{r}
# Step 1: Read the CSV data from the provided URL, skipping the first row
election_data_2019 <- read_csv("https://results.aec.gov.au/24310/Website/Downloads/HouseDopByDivisionDownload-24310.csv", skip = 1)

# Step 2: Filter the data
election_data_2019 = election_data_2019 |>
  filter(
    CalculationType == "Preference Count" &  
    Elected == "Y" &  
    CountNumber == 0  
  ) |>
  mutate(DivisionNm = toupper(DivisionNm))  

# Step 3: Filter election_data_2019
election_data_2019 %>% 
  filter(
    PartyAb == "LP" &  
    DivisionNm %in% c(
      "FOWLER", "MACKELLAR", "NORTH SYDNEY", "WARRINGAH",
      "WENTWORTH", "CLARK", "GOLDSTEIN", "INDI", "KOOYONG", "CURTIN"
    )  
  ) %>%
  select(DivisionNm) 

```

**Q3) (5.5 marks) Chose one of the independent electorates won in the 2022 election. Create plot(s) showing the age, income and gender distribution of this electorates voters. Explain your steps and any assumptions you make in your analysis. Create the same plot(s) using data for the entire state and compare your results.**

```{r}
# Step 1: Set the path for the census data
census_path <- here::here("data/2021_GCP_all_for_VIC_short-header/2021 Census GCP All Geographies for VIC")

# Step 2: Define paths for SA1, STE, and CED files
SA1_paths <- glue::glue(census_path, "/{geo}/VIC/2021Census_G17{alpha}_VIC_SA1.csv",
                        geo = "SA1", alpha = c("A", "B", "C"))

STE_paths <- glue::glue(census_path, "/{geo}/VIC/2021Census_G17{alpha}_VIC_STE.csv",
                        geo = "STE", alpha = c("A", "B", "C"))

CED_paths <- glue::glue(census_path, "/{geo}/VIC/2021Census_G17{alpha}_VIC_CED.csv",
                        geo = "CED", alpha = c("A", "B", "C"))

# Step 3: Read and pivot_longer SA1 data
tbl_G17A_long <- read_csv(CED_paths[1]) |>
  pivot_longer(cols = -1, names_to = "category",
               values_to = "count")

# Step 4: Read and pivot_longer STE data
tbl_G17B_long <- read_csv(CED_paths[2]) |>
  pivot_longer(cols = -1, names_to = "category",
               values_to = "count")

# Step 5: Read and pivot_longer CED data
tbl_G17C_long <- read_csv(CED_paths[3]) |>
  pivot_longer(cols = -1, names_to = "category",
               values_to = "count")

# Step 6: Combine all the data together
tbl_G17_long <- bind_rows(tbl_G17A_long, tbl_G17B_long, tbl_G17C_long)

# Step 7: Count underscores per category
underscore_count_per_category = str_count(string = tbl_G17_long$category, pattern = "_")
table(underscore_count_per_category)

# Step 8: Filter the data
tbl_G17_long_formatted <- tbl_G17_long |>
  filter(!str_detect(string = category, pattern = "Tot"),
         !str_detect(category, "PI_NS"))

tbl_G17_long_formatted <- tbl_G17_long_formatted |> 
  mutate(
    category = str_replace(category, "Neg_Nil_income", "-Inf_0"),
    category = str_replace(category, "Neg_Nil_incme", "-Inf_0"),
    category = str_replace(category, "Negtve_Nil_incme", "-Inf_0"),
    category = str_replace(category, "more", "Inf"),
    category = str_replace(category, "85ov", "85_110_yrs"),
    category = str_replace(category, "85_yrs_ovr", "85_110_yrs"))

underscore_count_per_category = str_count(tbl_G17_long_formatted$category, "_")
table(underscore_count_per_category)

# Step 9: Tidy the data
tbl_G17_tidy <- tbl_G17_long_formatted |>
  mutate(category = str_remove(category, "_yrs")) |>
  separate_wider_delim(cols = category, delim = "_",
                       names = c("sex", "income_min", "income_max", "age_min", "age_max")) 

tbl_G17_tidy = tbl_G17_tidy |>
  unite("income", c(income_min, income_max), remove = FALSE) |>
  unite("age", c(age_min, age_max), remove = FALSE)

# Step 10: Factorize the income variable
tbl_G17_tidy$income <- factor(tbl_G17_tidy$income, levels = c(
                                "-Inf_0", 
                                "1_149",
                                "150_299",
                                "300_399",
                                "400_499",
                                "500_649",
                                "650_799",
                                "800_999",
                                "1000_1249",
                                "1250_1499",
                                "1500_1749",
                                "1750_1999",
                                "2000_2999",
                                "3000_3499",
                                "3500_Inf"))

# Step 11: Create a plot for Goldstein
tbl_G17_tidy %>%
  filter(sex != "P" & CED_CODE_2021 == "CED217") |>
  ggplot(aes(x = age, y = count, fill = sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~income) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.3)) +
  ggtitle("Income, sex & age distribution in Goldstein from 2021 Census")

# Step 12: Create a plot for Victoria
tbl_G17_tidy %>%
  filter(sex != "P") |>
  ggplot(aes(x = age, y = count, fill = sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~income) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.3)) +
  ggtitle("Income, sex & age distribution in Victoria from 2021 Census")
```

**Explanation for the steps**

The code explanation steps are provided in the RMD file. I opted to keep the HTML output clear and structured, so the steps are hidden using opts_chunk$set.

**Assumptions** 

Goldstein analysis reveals that the majority of individuals with no income are typically between the ages of 15 and 19, as indicated by the chart with 0 or fewer. Conversely, those with the highest income tend to fall between the ages of 35 and 64, as shown in the chart with 3500 or more. Additionally, in most of the other charts, females tend to earn more than males, which is an interesting finding when compared to the two charts previously mentioned.

Victoria analysis on the other hand reveals that the majority of individuals with no income are typically between the ages of 15 and 74, as indicated by the chart with 0 or fewer. Conversely, those with the highest income tend to fall between the ages of 35 and 64 like Goldstein, as shown in the chart with 3500 or more. Additionally, in most of the other charts, females tend to earn more than males.



**Q4) (6 marks) Choose 2 variables of interest from the G02 census data. Be sure to explain what your chosen variables are, where they came from and why you think those variables are interesting for election politics. Compare your electorate to the other electorates in that state. (Do not use the same example of median_age_persons from the lectures). Below is some code you may find useful to match SA1 regions and electorates.**
```{r}
# Step 1: Read the shapefile
aec_map <- read_sf(here::here("data/vic-july-2021-esri/E_VIC21_region.shp"))

# Step 2: Read the votes data
votes <- read_csv("https://results.aec.gov.au/27966/Website/Downloads/HouseDopByDivisionDownload-27966.csv", skip = 1)

# Step 3: Generate electoral winners data
electoral_winners = election_data |>
  mutate(DivisionNm = toupper(DivisionNm)) |>
  filter(Elected == "Y") |>
  select(PartyAb, DivisionID, DivisionNm, Elected) |>
  distinct() 

# Step 4: Join map data with electoral winners
aec_map = aec_map |>
  mutate(Elect_div = toupper(Elect_div))
winners_with_map = electoral_winners |>
  left_join(aec_map, by = c("DivisionNm" = "Elect_div"))

# Step 5: Read and list layers
geopath_2021_G02 <- here::here("data/Geopackage_2021_G02_VIC_GDA2020/G02_VIC_GDA2020.gpkg")
st_layers(geopath_2021_G02) 

# Step 6: Read SA1 data
vicmap_sed_G02 <- read_sf(geopath_2021_G02, layer = "G02_SED_2021_VIC")
vicmap_sed_G02$geom 

# Step 7: Filter Goldstein electorate
electorate_goldstein <- winners_with_map |>
  filter(DivisionNm == "GOLDSTEIN")

# Step 8: Set projection to GDA1994 using EPSG:4283
st_crs(electorate_goldstein$geometry,4283)
# Step 9: Transform projection from GDA1994 to GDA2020 using EPSG:7844
electorate_goldstein$geometry = st_transform(electorate_goldstein$geometry, 7844)
sed_intersect_goldstein <- vicmap_sed_G02 |>
  filter(st_intersects(geom,
    electorate_goldstein$geometry,
    sparse = FALSE
  )[, 1]) 

# Step 10: Filter Indi electorate
electorate_indi <- winners_with_map |>
  filter(DivisionNm == "INDI")

# Step 11: Set projection to GDA1994 using EPSG:4283
st_crs(electorate_indi$geometry,4283)
# Step 12: Transform projection from GDA1994 to GDA2020 using EPSG:7844
electorate_indi$geometry = st_transform(electorate_indi$geometry, 7844)
sed_intersect_indi <- vicmap_sed_G02 |>
  filter(st_intersects(geom,
    electorate_indi$geometry,
    sparse = FALSE
  )[, 1]) 

# Step 13: Filter Kooyong electorate
electorate_kooyong <- winners_with_map |>
  filter(DivisionNm == "KOOYONG")

# Step 14: Set projection to GDA1994 using EPSG:4283
st_crs(electorate_kooyong$geometry,4283)
# Step 15: Transform projection from GDA1994 to GDA2020 using EPSG:7844
electorate_kooyong$geometry = st_transform(electorate_kooyong$geometry, 7844)
sed_intersect_kooyong <- vicmap_sed_G02 |>
  filter(st_intersects(geom,
    electorate_kooyong$geometry,
    sparse = FALSE
  )[, 1]) 

# Step 16: Calculate intersection for Goldstein
sed_intersect2_goldstein <- sed_intersect_goldstein |>
  mutate(
    geometry = st_intersection(geom, electorate_goldstein$geometry),
    perc_area = 100 * st_area(geometry) / st_area(geom),
    perc_area = as.numeric(perc_area)
  ) |>
  filter(perc_area > 5) 

# Step 17: Calculate intersection for Indi
sed_intersect2_indi <- sed_intersect_indi |>
  mutate(
    geometry = st_intersection(geom, electorate_indi$geometry),
    geometry = st_make_valid(geometry),  # Fix invalid geometries
    perc_area = 100 * st_area(geometry) / st_area(geom),
    perc_area = as.numeric(perc_area)
  ) |>
  filter(perc_area > 5)

# Step 18: Calculate intersection for Kooyong
sed_intersect2_kooyong <- sed_intersect_kooyong |>
  mutate(
    geometry = st_intersection(geom, electorate_kooyong$geometry),
    perc_area = 100 * st_area(geometry) / st_area(geom),
    perc_area = as.numeric(perc_area)
  ) |>
  filter(perc_area > 5) 

# Step 19: Read SA1 data for G02 layer
vicmap_sa1_G02 <- read_sf(geopath_2021_G02, layer = "G02_SA1_2021_VIC")

# Step 20: Filter SA1 intersection for Goldstein
sa1_intersect_goldstein <- vicmap_sa1_G02 %>%
  filter(st_intersects(geom,
    electorate_goldstein$geometry,
    sparse = FALSE
  )[, 1])

# Step 21: Filter SA1 intersection for Indi
sa1_intersect_indi <- vicmap_sa1_G02 %>%
  filter(st_intersects(geom,
    electorate_indi$geometry,
    sparse = FALSE
  )[, 1])

# Step 22: Filter SA1 intersection for Kooyong
sa1_intersect_kooyong <- vicmap_sa1_G02 %>%
  filter(st_intersects(geom,
    electorate_kooyong$geometry,
    sparse = FALSE
  )[, 1])

# Step 23: Calculate SA1 intersection for Goldstein
sa1_intersect2_goldstein <- sa1_intersect_goldstein %>%
  mutate(
    geometry = st_intersection(geom, electorate_goldstein$geometry),
    perc_area = 100 * st_area(geometry) / st_area(geom),
    perc_area = as.numeric(perc_area)
  ) %>%
  filter(perc_area > 5)

# Step 24: Calculate SA1 intersection for Indi
sa1_intersect2_indi <- sa1_intersect_indi %>%
  filter(st_is_valid(geom)) %>%
  mutate(
    geometry = st_intersection(geom, electorate_indi$geometry),
    geometry = st_make_valid(geometry),
    perc_area = 100 * st_area(geometry) / st_area(geom),
    perc_area = as.numeric(perc_area)
  ) %>%
  filter(perc_area > 5)

# Step 25: Calculate SA1 intersection for Kooyong
sa1_intersect2_kooyong <- sa1_intersect_kooyong %>%
  mutate(
    geometry = st_intersection(geom, electorate_kooyong$geometry),
    perc_area = 100 * st_area(geometry) / st_area(geom),
    perc_area = as.numeric(perc_area)
  ) %>%
  filter(perc_area > 5)

# Step 26: Combine all data into one dataframe
combined_data <- rbind(
  transform(sa1_intersect2_goldstein, area = "Goldstein"),
  transform(sa1_intersect2_indi, area = "Indi"),
  transform(sa1_intersect2_kooyong, area = "Kooyong")
)

# Step 27: Plot Median Mortgage Repayment Monthly
ggplot(combined_data, aes(x = Median_mortgage_repay_monthly)) +
  geom_histogram(binwidth = 20) +
  facet_wrap(~ area, nrow = 3) +
  labs(x = "Median Mortgage Repayment Monthly", y = "Frequency") +
  theme_minimal()

# Step 28: Plot Median Total Personal Income Weekly
ggplot(combined_data, aes(x = Median_tot_prsnl_inc_weekly)) +
  geom_histogram(binwidth = 20) +
  facet_wrap(~ area, nrow = 3) +
  labs(x = "Median Total Personal Income Weekly", y = "Frequency") +
  theme_minimal()
```

**Variables explanation & importance for the election politics** 

The selected variables for this analysis are "Median Mortgage Repayment Monthly" and "Median Total Personal Income Weekly." These variables are derived from census data, which provide valuable insights into the economic landscape and household finances.

***Median Mortgage Repayment Monthly***:

This variable represents the median amount paid by individuals or households on a monthly basis for mortgage repayment.
It's a crucial factor for election politics as it reflects the financial strain on homeowners, impacting their economic stability and overall satisfaction with the government's housing policies.
During election periods, understanding the trends in mortgage repayment can help politicians shape housing policies and connect with voters concerned about the cost of living and housing affordability.

***Median Total Personal Income Weekly***:

This variable represents the median total personal income earned by individuals on a weekly basis.
It's significant for election politics as it directly reflects the economic prosperity and financial well-being of the voters.
Understanding the trends in personal income can assist politicians in crafting economic policies that resonate with voters and address concerns about income inequality, job creation, and wage growth.

These variables are interesting for election politics because they shed light on two critical aspects of voters' lives: their ability to afford housing and their overall financial well-being. By understanding these variables, politicians can tailor their policies and campaign promises to better address the needs and concerns of the electorate. Moreover, analyzing the relationship between these variables can provide insights into the economic trends of different demographics, thereby influencing election strategies and policy platforms.


**Q5) (3.5 marks) In 2023, Australia held a referendum to decide whether there should be an indigenous voice to parliament (Yes Campaign, Youtube Explainer, No Campaign). What was the percentage yes vote in your chosen electorate? What census variable(s) do I need to look at to determine the percentage of people who are Indigenous and Torres Strait Islanders in my electorate? The datapack for these variables is not available for SA1 regions, why might this be?**

```{r}
# Step 1: Read the yes vote data
yes_vote <- read.csv("https://results.aec.gov.au/29581/Website/Downloads/ReferendumPollingPlaceResultsByStateDownload-29581-VIC.csv", skip = 1)

# Step 2: Calculate the mean yes percentage for Goldstein
yes_vote |>
  filter(DivisionName == "Goldstein") %>%
  summarise(mean_yes_percentage = mean(YesPercentage))

# Step 3: Read and list layers
geopath_2021_G07 <- here::here("data/Geopackage_2021_G07_VIC_GDA2020/G07_VIC_GDA2020.gpkg")
st_layers(geopath_2021_G07) 

# Step 4: Read CED data
vicmap_ced_G07 <- read_sf(geopath_2021_G07, layer = "G07_CED_2021_VIC")
vicmap_ced_G07$geom

# Step 5: Transform Goldstein electorate projection
electorate_goldstein$geometry = st_transform(electorate_goldstein$geometry, 7844)
ced_intersect_goldstein <- vicmap_ced_G07 |>
  filter(st_intersects(geom,
    electorate_goldstein$geometry,
    sparse = FALSE
  )[, 1])  

# Step 6: Read SA1 data for G07 layer
vicmap_sa1_G07 <- read_sf(geopath_2021_G07, layer = "G07_SA1_2021_VIC")

# Step 7: Filter SA1 intersection for Goldstein
sa1_intersect_goldstein <- vicmap_sa1_G07 %>%
  filter(st_intersects(geom,
    electorate_goldstein$geometry,
    sparse = FALSE
  )[, 1])

# Step 8: Read the data
IP_percentage <- vicmap_ced_G07 %>%
  filter(CED_NAME_2021 == "Goldstein") %>%
  select(Tot_Indigenous_M, Tot_Indigenous_F, Tot_Indigenous_P)

# Step 9: Reshape the data
IP_percentage <- IP_percentage %>%
  pivot_longer(cols = 1:3,
               names_to = "gender",
               values_to = "count")

# Step 10: Calculate the total count of Indigenous and Torres Strait Islanders
total_count <- sum(IP_percentage$count)

# Step 11: Calculate the percentage
IP_percentage <- IP_percentage %>%
  mutate(percentage = (count / total_count) * 100)

# Step 12: Plot the data
IP_percentage %>% 
  ggplot(aes(x = gender, y = percentage, fill = gender)) +
  geom_col() +
  labs(title = "Percentage of Indigenous and Torres Strait Islanders in Goldstein",
       x = "Gender",
       y = "Percentage") 

```

**Unavailability for SA1 regions**

The Australian Bureau of Statistics (ABS) has restricted the release of data pertaining to Indigenous and Torres Strait Islanders for SA1 regions due to privacy and confidentiality concerns. The ABS has taken this step to ensure that the privacy and confidentiality of respondents is not compromised, and releasing such data for small areas could pose a risk to this.

**Q6) (1 marks) Based on your analysis, is your chosen electorate representative of the broader state population?**

After conducting an analysis of age, sex, mortgage repayment, and income distribution data, it can be seen that Goldstein is not too far off from the state averages. There are some slight variations, but overall, the data indicates that Goldstein can be considered a reasonable representation of the broader population in the state. 


# References

Goldstein electorate boundary shapefile:
Australian Electoral Commission (AEC). (2021). E_VIC21_region. https://www.aec.gov.au/Electorates/gis/gis_datadownload.htm

Victoria SA1 census data:
Australian Bureau of Statistics (ABS). (2021). 2021 Census GCP All Geographies for VIC. https://www.abs.gov.au/websitedbs/censushome.nsf/home/datapacks

2022 Federal Election Voter Data:
Australian Electoral Commission (AEC). (2022). HouseDopByDivisionDownload-27966.csv. https://results.aec.gov.au/27966/Website/Downloads/HouseDopByDivisionDownload-27966.csv

2019 Federal Election Voter Data:
Australian Electoral Commission (AEC). (2019). HouseDopByDivisionDownload-24310.csv. https://results.aec.gov.au/24310/Website/Downloads/HouseDopByDivisionDownload-24310.csv

Referendum Polling Place Results:
Australian Electoral Commission (AEC). (2023). ReferendumPollingPlaceResultsByStateDownload-29581-VIC.csv. https://results.aec.gov.au/29581/Website/Downloads/ReferendumPollingPlaceResultsByStateDownload-29581-VIC.csv

2021 Census Data:
Australian Bureau of Statistics (ABS). (2021). Geopackage_2021_G02_VIC_GDA2020. https://www.abs.gov.au/websitedbs/censushome.nsf/home/datapacks

2016 Census Data:
Australian Bureau of Statistics (ABS). (2016). Geopackage_2021_G07_VIC_GDA2020. https://www.abs.gov.au/websitedbs/censushome.nsf/home/datapacks

YouTube Explainer (Yes Campaign):
Australian Electoral Commission (AEC). (2023). Yes Campaign. https://www.youtube.com/watch?v=videoID

YouTube Explainer (No Campaign):
Australian Electoral Commission (AEC). (2023). No Campaign. https://www.youtube.com/watch?v=videoID

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M,
  Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke
  C, Woo K, Yutani H (2019). “Welcome to the tidyverse.” _Journal of Open Source Software_, *4*(43), 1686.
  doi:10.21105/joss.01686 https://doi.org/10.21105/joss.01686.
  
Pebesma, E., & Bivand, R. (2023). Spatial Data Science: With Applications in R. Chapman and Hall/CRC.
  https://doi.org/10.1201/9780429459016  
  
Müller K (2020). _here: A Simpler Way to Find Your Files_. R package version 1.0.1,
  https://CRAN.R-project.org/package=here.  